\chapter{Expansion and Civil War in America}

We look at the entire middle of the 19th century in America in this chapter,
starting with the Mexican-American War and going straight through to the end of Reconstruction.

\section{War With Mexico}

\subsection*{James K.\ Polk}

The election of 1844 resulted in Democrat \textbf{James K.\ Polk}, ``Young Hickory'', defeating the Whig Henry Clay,
following Clay's third and final attempt to take the office.
Polk, who was governor of Tennessee, had been a dark horse who got the nomination over Lewis Cass and Martin Van Buren.

Polk's Vice President was \textbf{George M.\ Dallas}.
Robert Walker, Secretary of the Treasury, passed the Walker Tariff, which helped a border dispute in Oregon;
this resulted in Polk acquiring some of the Oregon territory for the US\@.
Secretary of the Navy George Bancroft established Annapolis Naval Academy,
and Polk established the Smithsonian.
As president, he also created the Department of the Interior.

\subsection*{Revolution in Texas}

In 1836, American settlers in Texas (a part of Mexico) decided they didn't want to be Mexican any more.
So they declared independence and created the \textbf{Republic of Texas}.

Mexican President \textbf{Antonio Lopez de Santa Anna} decided to take Texas back.
His general Jose de Urrea led the Goliad campaign up the coast, killing almost all troops he encountered.

Santa Anna himself led a force that marched to San Antonio, where he met a Texan garrison at the \textbf{Alamo}.
Before the Battle of the Alamo, the commander William Travis acknowledged certain death.
He drew a line in the sand and said that anyone who wanted to could cross the line and fight with him.
Everyone crossed the line except for Moses Rose, who ran away.
Defenders at the Alamo included \textbf{Davy Crockett} and Jim Bowie.
Travis responded to Santa Anna's request for surrender by firing cannons at him.
The garrison of about 200 men died fighting Santa Anna, in a battle that is remembered to this day.

The Texans, under \textbf{Sam Houston} (first president of Republic of Texas),
came back to ambush Santa Anna's forces at \textbf{San Jacinto}.
The battle didn't even last 20 minutes because the Texans caught the Mexicans on a siesta.
The Texans also had some impressive cannons called the ``Twin Sisters''.
Battle cries included ``Remember Goliad'' and ``Remember the Alamo!''.
The battle was a decisive victory for the Texans,
and it resulted in the capture of Santa Anna and the signing of Treaties of Velasco.

\subsection*{Course of War}

In 1845, the US annexed Texas, which Mexico considered part of Mexico,
even though Texas had a Revolution in 1836 against Mexico.
Precipitated by the Thornton Affair, a battle along the Rio Grande,
the \textbf{Mexican-American War} lasted from 1846 to 1848.
The US quickly took New Mexico and California.

The war featured some notable fighting groups.
St\ Patrick's Battalion was a group of Irish Catholic soldier who defected and were subsequently hanged.
The Mormon Battalion, the only religious military unit in US history, was led by Philip Saint George Cooke.

At the Battle of Chapultapec, US Marines stormed a castle on a hill, the site of a Mexican military academy.
They won, and in the process, they killed the six ``Boy Heroes'' (Los Ni\~nos H\'eroes),
Mexican military cadets who wouldn't fall back when ordered by General Bravo.

\textbf{Zachary Taylor} led land forces to win at battles at Monterrey and Buena Vista.
Winfield Scott and Matthew Perry ran the first successful American amphibious landing at \textbf{Veracruz}.
They besieged the city and took it.

In 1846, Californian settlers under William B.\ Ide captured Sonoma and General Mariano Vallejo.
This initiated the \textbf{Bear Flag Revolt}, and John Fremont (the Pathfinder) and Robert Stockton became leaders.
The revolt was put down by Stephen Kearny during his conquest of California during the war.

The war was ended by the \textbf{Treaty of Guadalupe-Hidalgo}.
Nicholas Trist and John Slidell were negotiators at the treaty,
but Trist's disobedience of Polk's instructions made Polk mad at him.
Trist had been ordered to offer \$30 million to get Baja California and more territory in the South.
When Trist finished with the treaty, he had taken less land, and he got money from Mexico,
and was promptly fired on returning to Washington.

\section{American Frontier}

\subsection*{Outlaws of the Wild West}
% Jesse James
% Wild Bill Hickok - Dead man's hand
% Doc Holliday
% Wyatt Earp
% Billy the Kid

\subsection*{Pacific Coast}
% California Gold Rush

\section{Last of the Whigs}
% Taylor
% Fillmore
% Pierce

\section{Civil War}

\subsection*{Abolitionism}
% John Brown
% Harper's Ferry
% Free Soil - 1848 van Buren
% Knights of the Golden Circle

\subsection*{Prelude}
% Buchanan
% Lewis Cass
% John Brown - Pottawatomie Creek, Harper's Ferry
% Compromise of 1850
% Crittenden compromise
% Kansas-Nebraska Act - Bleeding Kansas
% Caning Sumner

\subsection*{Battles and Commanders}

\section{Reconstruction}
% Wade Davis

\subsection*{Andrew Johnson}
% Edwin Stanton
% Tenure of Office Act

\subsection*{Grant and Scandal}
% Whiskey Ring

